---
title: "An Introdution to Geospatial Data Structures"
author: 'Yibo Liang, Pippa Lin, '
date: December 3, 2023
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

## 1.1 The "sf" Package

The R-package we will be using for this demonstration is the "sf" package. It adopts the "Simple Features" formal standard, outlining a storage and access model for spatial geometries such as points, lines, and polygons. A geometry is considered simple when it comprises points connected by straight line segments and avoids self-intersections.

```{r, results='hide'}
library(sf)
library(tidyverse)
library(tidymodels)
library(ggplot2)
library(dplyr)
library(ggmap)
library(leaflet) 
```

## 1.2 Basic Spatial Structures in "sf"

Some of the basic spatial objects that are in the "sf" package for dealing with vector spatial data include:

-   points

-   lines

-   polygons

### 1.2.1 Creating a "sf" object

Let's create an example "sf" line object. First, we will define the lines themselves with `st_linestring()`

```{r}
lnstr_sfg1 <- st_linestring(matrix(runif(10), ncol=2)) # a line with 10 points
lnstr_sfg2 <- st_linestring(matrix(runif(10), ncol=2)) # ncol = 2 specifies the dimension
```

Next, we will combine the two lines into a single feature before transforming it into a single "sf" object.

```{r}
(lnstr_sfc <- st_sfc(lnstr_sfg1, lnstr_sfg2)) # just one feature here
```

Now that we have 2 lines combined into a single feature, we haven't actually assigned any meaning to the 2 lines yet. Let's give them some attributes! Let's pretend that these two lines are a highway and we they each have an attribute \`car_per_hour\` describing the number of cars that travels through them every hour.

```{r}
dfr <- data.frame(id = c("hwy1", "hwy2"),
                  cars_per_hour = c(78, 22)) 
(lnstr_sf <- st_sf(dfr , lnstr_sfc)) # 2 lines each with an attribute cars_per_hour
```

Now we have a data-frame that associates an attribute to a spatial structure rather than another data point!

## 2.1 Geospatial Dataframe

```{r}
us_species <- read_csv("data/dat_species.csv")
us_species <- us_species %>% filter(group == "Mammals"| group == "Reptiles" | group == "Amphibians" | group == "Amphibians-Frogs")
us_geo <- read_csv("data/dat_spatial.csv")
us_geo <- us_geo %>% 
  filter(name != "Hawaii" & name != "Alaska")
```

```{r}
colnames(us_geo)[1] <- "state"
us_geo_merged <- merge(us_geo,us_species,by="state", all=T)
names(us_geo_merged) 
```

```{r}
us_geo_merged <- us_geo_merged[is.na(us_geo_merged$geometry) == FALSE, ]

# Create an sf object (SQ) from the WKT geometries
SQ <- st_as_sf(us_geo_merged, wkt = "geometry")
SQ1 <- st_as_sf(us_geo, wkt = "geometry")
```

```{r}
# Check the structure of SQ
str(SQ)
unique(us_geo_merged$sciname)
```

```{r}
SQ_f <- SQ %>%
  filter(sciname == "Hydrochoerus hydrochaeris") %>% 
  filter(year <= 2015) %>% 
  group_by(state) %>% 
  summarise(occurrence = sum(occurrence))

  # Create a ggplot object
p <- ggplot() +
  # Add the filtered sf object as a layer
  geom_sf(data = SQ1["state"]) +
  geom_sf(data = SQ_f, aes(fill = occurrence)) +
        
  # Customize the plot as needed
  theme_minimal() +
    labs(title = "Occurrences Plot")

p
```
